The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2018-12-16 pkg/partykit/inst/guideliketest/draft/draft.Rnw
restructuring of chapter on unbiased recursive partitioning and on the unifying framework; selection of presented results; further revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2393 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c81d78a543e39934aa64777e9b14cacec5c1d449

%% packages
\usepackage{amstext,amsfonts,amsmath,amssymb,bm,thumbpdf,lmodern}

%% need no \usepackage{Sweave}
\SweaveOpts{engine = R, concordance = FALSE, eps = FALSE, keep.source = TRUE, eval = TRUE}

@@

<<preliminaries, echo=FALSE, results=hide>>=
# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/draft/")
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("partykit")
library("Formula")
library("parallel")
library("lattice")
source("../guidelike.R")
source("../tests.R")
load("../sim/simres20180422_1step.rda")
sim1step <- simres
#load("../sim/simres20180422_2step.rda")
#sim2step <- simres
rm(simres)
@

\title{Comparing Testing Strategies of Unbiased Tree Algorithms}
%\Shorttitle{Comparing Different Testing Approaches in Tree Algorithms}

@@

\Abstract{Comparing Testing Strategies of Unbiased Tree Algorithms}
\Keywords{classification and regression trees, independence tests, recursive partitioning}

@@

\section{Introduction}
In many situations fitting one global model to a data set can be very challenging, especially if this data contains lots of different features with high/extreme variations or complex interactions. Therefore, separating the data into more homogeneous subgroups based on (a set of) covariates first can simplify the task and fitting a local model to each of the resulting subgroups often leads to better results. This separation of the data can be done by applying a tree algorithm. Many different approaches and algorithms have been developed in this research field. While they all follow the idea of splitting the data in a way such that some objective function is minimized/maximized, they differ in their approach to the crucial decision that has to made before each split: In which variable and at which break point should a split be performed. (Early tree algorithms rely on exhaustive search procedures to find the best split point and split variable in one step. However, it has been shown that this is not only computatitionaly expensive but also biased towards split variables with many possible split points. Therefore, succeeding algorithms select a split variable in a first step and then search for the best split point within this variable only in a second step. 
In this paper the focus will be put on the task of selecting the (right/best) split variable. 
In particular the strategy of the GUIDE \citep{Loh:2002} algorithm is compared to the one of the CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} and the MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008}. Moreover, adapted versions of these algorithms are evaluated as well in order to compare effects of specific differences separately. 
After introducing the setting for the simulation study in Section~\ref{sec:settings} and a general scheme on how to build a tree in Section~\ref{sec:tree}, the three main differences will be discussed in Sections~\ref{sec:scores}, \ref{sec:categorization} and \ref{sec:binarization}.

\section{Simulation Settings}
\label{sec:settings}
\subsection{Data Generating Process}
The different approaches are evaluated on generated data sets. Each data set consists of a response variable, one regressor, one split variable and nine noise variables. In particular, the variables listed in Table~\ref{tab:variables} together with the corresponding distributions are included in the data generating process.

\vspace{0.2cm}

\begin{table}
\begin{center}
\begin{tabular}{l l l}
Name                    & Notation                   & Specification \\
\hline
response                & $Y$                        & $\mathcal{N}(\mu(X, Z_1), \sigma^2)$ \\
mean of $Y$             & $\mu$                      & linear function of $X$ and $Z_1$ \\
variance of $Y$         & $\sigma^2$                 & 1\\
regressor               & $X$                        & $\mathcal{U}([-1,1])$  or $(-1)^{\mathcal{B}(0.5)}$ \\
true split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
coefficients            & $\beta_0$, $\beta_1$       & functions of $Z_1$\\
true split point        & $\xi$                      & $\in \{0, 0.2, 0.5, 0.8\}$\\
effect size             & $\delta$                   & $\in \{0, 0.25, 0.5, 0.75, 1\}$
\end{tabular}
\caption[Table caption text]{Variables included in the data generating process.}
\label{tab:variables}
\end{center}
\end{table}

\vspace{0.4cm}

\noindent The location parameter $\mu$ of the normally distributed response variable $Y$ depends linearly on the regressor variable $X$ where the intercept $\beta_0$ and the slope parameter $\beta_1$ of this linear equation (can) depend on the true split variable $Z_1$ as explained below.
\begin{center}
$\mu(X, Z_1) = \beta_0(Z_1) + \beta_1(Z_1) \cdot X$
\end{center}

\noindent For the coefficients $\beta_0$ and $\beta_1$ the following three different scenarios are considered:
\begin{enumerate}
\item The intercept $\beta_0$ varies depending on $Z_1$ while $\beta_1$ is fixed (at 1).
\item The slope coefficient $\beta_1$ varies depending on $Z_1$ while $\beta_0$ is fixed (at 0).
\item Both coefficients $\beta_0$ and $\beta_1$ vary depending on $Z_1$.
\end{enumerate}

\noindent The varying coefficient(s) $\beta_i$, $i\in\{0,1\}$, can either be binary, e.g. a step function, or a continuous function. In the binary case the following definition is applied
\begin{align*}
\beta_{i}(Z_1) = \begin{cases}
-\delta \cdot (-1)^{i}  \quad \text{if } Z_1 < \xi \\
+\delta \cdot (-1)^{i}  \quad \text{if } Z_1 \geq \xi
\end{cases}
\end{align*}
while in the continuous case a simple linear function is considered in the form of
$$\beta_{i}(Z_1) = Z_1 \cdot \delta \cdot (-1)^{i}.$$
In that way, the type of variation is the same for $\beta_0$ and $\beta_1$, however, in oposite directions.

%\noindent Depending on whether the coefficients are binary or continuous, in the case of only one coefficient varying it is defined by 
%\begin{align*}
%\beta_{0/1}(Z_1) = \begin{cases}
%+\delta \qquad  \quad \text{if } Z_1 < \xi \text{ and }\texttt{binary\_beta=TRUE}\\
%-\delta \qquad  \quad \text{if } Z_1 \geq \xi \text{ and }\texttt{binary\_beta=TRUE}\\
%Z_1 \cdot \delta \qquad \text{if } \texttt{binary\_beta=FALSE}
%\end{cases}
%\end{align*}

%\noindent while in the case of both coefficients varying they are defined by:
%\begin{align*}
%\beta_{0}(Z_1) = \begin{cases}
%-\delta \qquad  \quad \text{if } Z_1 < \xi \text{ and }\texttt{binary\_beta=TRUE}\\
%+\delta \qquad  \quad \text{if } Z_1 \geq \xi \text{ and }\texttt{binary\_beta=TRUE}\\
%Z_1 \cdot \delta \qquad \text{if } \texttt{binary\_beta=FALSE}
%\end{cases}
%\end{align*}
%\begin{align*}
%\beta_{1}(Z_1) = \begin{cases}
%+\delta \qquad  \quad \text{if } Z_1 < \xi \text{ and }\texttt{binary\_beta=TRUE}\\
%-\delta \qquad  \quad \text{if } Z_1 \geq \xi \text{ and }\texttt{binary\_beta=TRUE}\\
%- Z_1 \cdot  \delta \hspace{0.5cm} \text{if } \texttt{binary\_beta=FALSE}
%\end{cases}
%\end{align*}

%\newpage
\begin{figure}%[t!]
\setkeys{Gin}{width=1.0\linewidth}
<<dgp_bin_beta_cont_reg, fig=TRUE, echo=FALSE>>=
set.seed(7)

nobs <- 300
sigma <- 0.5
xi <- 0
delta <- 1
beta0 <- 0
beta1 <- 1

@@

par(mar = c(4,4,2,1))
plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", col = "blue", main = "vary beta_0")
points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), col = "red")
legend("top", c("z1 < 0", "z1>=0"), col = c("blue", "red"), pch = c(19), bty = "n")
plot(x = d$z1, y = d$beta0, ylab = "beta0", xlab = "z1", main = "vary beta_0")
plot(x = d$z1, y = d$beta1, ylab = "beta1", xlab = "z1", main = "vary beta_0")

@@

par(mar = c(4,4,2,1))
plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", col = "blue", main = "vary beta_1")
points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), col = "red")
legend("top", c("z1 < 0", "z1>=0"), col = c("blue", "red"), pch = c(19), bty = "n")
plot(x = d$z1, y = d$beta0, ylab = "beta0", xlab = "z1", main = "vary beta_1")
plot(x = d$z1, y = d$beta1, ylab = "beta1", xlab = "z1", main = "vary beta_1")

>>>>

%% packages
\usepackage{amstext,amsfonts,amsmath,amssymb,bm,thumbpdf,lmodern,hyperref}
\usepackage[all]{hypcap}

%% need no \usepackage{Sweave} ?
%\usepackage{Sweave}
\SweaveOpts{engine = R, concordance = FALSE, eps = FALSE, keep.source = TRUE, eval = TRUE}

@@

<<preliminaries, echo=FALSE, results=hide>>=
# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/draft/")
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("partykit")
library("Formula")
library("parallel")
library("lattice")
library("scales")
# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/draft/")
source("../guidelike.R")
source("../tests.R")
#load("../sim/simres20180422_1step.rda")
load("../sim/simres20181122_stump.rda")
sim_stump <- simres$res
#load("../sim/simres20180422_2step.rda")
#sim2step <- simres
rm(simres)
load("../sim/simres20181122_tree.rda")
sim_tree <- simres$res
rm(simres)

## HCL palette
pal <- hcl(c(140, 80, 260, 50, 10), 300, 60)
names(pal) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_full")

@

\title{The Power of Unbiased Recursive Partitioning:\\
A Unifying View of CTree, MOB, and GUIDE}
\Shorttitle{The Power of Unbiased Recursive Partitioning}
%Comparing Testing Strategies of Unbiased Tree Algorithms

@@

\Abstract{
A core step of every algorithm for learning regression trees is the decision if, 
how, and where to split the underlying data - or, in other words, the selection 
of the best splitting variable from the available covariates and the corresponding 
split point. Early tree algorithms (e.g., AID, CART) employ greedy search strategies 
that directly compare all possible split points in all available covariates. However, 
subsequent research showed that this is biased towards selection of covariates with 
more potential split points. Therefore, unbiased recursive partitioning algorithms 
have been suggested (e.g., QUEST, GUIDE, CTree, MOB) that first select the covariate 
based on statistical inference using p~-~values that are appropriately adjusted for 
the possible split points. In a second step a split point optimizing some objective 
function is selected in the chosen split variable. However, different unbiased tree 
algorithms employ different inference frameworks for computing these p~-~values and 
their relative advantages or disadvantages are not well understood, yet.

Therefore, three different approaches are considered here and embedded into a common 
modeling framework with special emphasis to linear model trees: classical categorical 
association tests (GUIDE), conditional inference (CTree), parameter instability 
tests (MOB). It is assessed how different building blocks affect the power of the 
tree algorithms to select the appropriate covariates for splitting: residuals vs. full 
model scores, binarization of residuals/scores at zero, binning of covariates, 
conditional vs. unconditional approximations of the null distribution.
}
\Keywords{classification and regression trees, independence tests, recursive partitioning}

@@

\section{Introduction}\label{sec:introduction}
In many situations fitting one global model to a data set can be very challenging, 
especially if this data contains lots of different features with high/extreme variations 
or complex interactions. Therefore, separating the data into more homogeneous subgroups 
based on (a set of) covariates first can simplify the task and fitting a local model to 
each of the resulting subgroups often leads to better results. This separation of the 
data can be done by applying a tree algorithm. Many different algorithms have been developed
in this research field. While they all follow the general idea of splitting the data such 
that some objective function is minimized/maximized, they differ in their 
specific/particular approaches to selecting a split variable and the corresponding split point. 
Some of the first tree algorithms (e.g., AID \citealp{Morgan+Sonquist:1963}; 
CART \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
the best split point and split variable in one step by directly comparing all possible split
points in all possible split variables. However, it has been shown that this is not only
computationally expensive but also biased towards split variables with many possible split
points \citep{Doyle:1973, Kim+Loh:2001}. 
Therefore, selecting a split variable in a first step and then searching for the 
best split point only within this variable in a separated second step is a more 
promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
CHAID \citep{Kass:1980}, GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
For the first step of selecting a split variable they all follow the idea of 
choosing the covariate which shows the highest association to/influence on the response variable 
based on a statistical test. While QUEST, CHAID and GUIDE employ statistical significance
tests for contingency tables, CTree applies permutation tests in a conditional inference
framework and MOB uses fluctuation tests based on asymptotic theory. 
In this paper the focus will be put on this first step, i.e. the task of selecting 
the right/best split variable. In particular, the approach of the GUIDE algorithm is
compared to the one of the CTree algorithm and the MOB algorithm. For this purpose, a 
unifying framework for unbiased model based tree algorithms is presented such that each 
of the three algorithms can be obtained by a specific combination of the available building 
blocks. In that way, the effects of each of these characteristic features of the GUIDE,
CTree and MOB algorithm can be investigated separately by varying the switches in
the tool box provided by this general framework.

In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree is given. 
Then, after explaining the investigated testing strategies more in detail in 
Section~\ref{sec:strategies} the unifying framework for model based unbiased
recursive partitioning is presented in Section~\ref{sec:unify}.
The setting for the simulation study is introduced in Section~\ref{sec:simulation}
and the results are discussed in Section~\ref{sec:results}.

\section{Unbiased Recursive Partitioning}
\label{sec:urp}
The basic idea of building a regression tree model is to partition the data into smaller and more 
homogeneous subgroups based on (a set of) covariates. Various different tree algorithms 
have been developed, all following the same general structure:
\begin{enumerate}
\item For the whole underlying data set in the current node some form of discrepancy 
measure is calculated. 
%This can for example be the sum of deviations from a typical or average value of the 
%data set or it can be based on a fitted model.
\item Among all available covariates one is selected as split variable and a value 
within the range/measurable space of this variable is chosen as split point such that 
after splitting the data the discrepancy measure is minimal within the resulting subgroups.
\item Steps 1 and 2 are repeated within each new subgroup until either a stopping criterion
is reached or the subgroup consists of only one or all equal observations. (In the second
case the resulting tree is of maximal size and can be reduced by (post-)pruning.)
\end{enumerate}

The employed discrepancy measure can for example be the sum of deviations from a typical 
or average value, but it can also be constructed based on a statistical model. For instance,
different types of residuals or their signs can be applied (see e.g., \cite{Loh:2002}) as well as
rank sums or logrank scores as in \cite{Hothorn+Hornik+Wiel:2006}.
(In the simulation study of this paper trees with linear models fitted in each node are 
investigated with the residual sum of squares (RSS) employed as discrepancy measure. However, 
the discussed methodology and the presented results also apply for a wide range of other 
model types.)

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
GUIDE first select the split variable and then the split point in a separate step. 
For the first step of selecting a split variable they all apply statistical (independence) tests
following the same general strategy: 
Based on the discrepancy measure an $n \times k$ - matrix $D$ is obtained giving information
on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
current node and $k \in \mathbb{N}$.
Using this matrix, the algorithm-specific test statistic is calculated for each possible split 
variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, evaluating the dependency
of $Z_j$ on $D$. As the variables are usually not all measured at the same scale the resulting
values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
necessary to calculate the corresponding p~-~values in order to allow for an unbiased
variable selection on this standardized/unified scale. Then the variable corresponding to the smallest
p~-~value is selected as split variable as it has shown to be the variable with the strongest
influence on the goodness of fit of the model.
% Hinweis: pruning-Strategien
%Once the split variable is chosen the corresponding split point is then selected such that it 
%minimizes the discrepancy measure.

While this basic approach is the same for the GUIDE, the CTree and the MOB algorithm they differ 
in their strategies on how to calculate test statistics. In order to point out these specific 
characteristics in Section~\ref{sec:unify} the strategies of the three tree algorithms are first 
explained in detail in Section~\ref{sec:strategies}.

%To emphasize the differences between these three algorithms in how they select split variables,
%their testing strategies are explained in detail in Section~\ref{sec:strategies} 
%using the example of tree models with a linear model fitted in each node as discussed
%in the simulation study. Again, this is done only for means of simplification and the same 
%methodology can be applied for a wide variety of models.

\subsection{Testing strategies}\label{sec:strategies}

\subsubsection{CTree}\label{sec:ctree}
The CTree algorithm is based on the idea of providing non-parametric regression tree models in 
a conditional inference framework by applying permutation tests. In its basic form linear test 
statistics are applied to measure the dependency between a covariate and the response variable. 
Optionally, maximally selected statistics can be employed as well. By allowing for transformations 
of the covariates and the application of so~-~called influence functions on the response variable 
within the test statistic a model-based approach is also possible. For example, the influence
function can calculate a model-based measure of goodness of fit as a discrepancy measure 
for each response value (in a permutation symmetric way) such that the test statistic evaluates 
the dependency of a covariate on the goodness of fit.
Following \cite{Strasser+Weber:1999} the conditional expectation and covariance of the test 
statistic given all permutations of the response variable can be calculated and used to standardize 
the test statistic. In order to map multivariate test statistics into the real line two options are 
given: either the maximum of the absolute values of the standardized statistic or a quadratic form 
is considered. The resulting values can then be transformed to the scale of p~-~values where they 
can be compared in an unbiased way to select a split variable.
In the CTree algorithm the number of possible breakpoints is fixed as splits can only be made at
observed values or optionally at an average value lying between two observations. To select the
best split point two-sample linear statistics of the same structure as in the independence test 
are evaluated for all possible subsets of the sample space of the selected split variable.

\subsubsection{MOB}\label{sec:mob}
While CTree offers to apply model-based discrepancy measures as one possible choice, the MOB algorithm 
was developed with the aim of embedding parametric models into the framework of regression trees
and is hence purely intended for a model-based approach.

An objective function $\psi(Y,X,\theta)$ depending on observations $(Y,X)$, with a response $Y$ and 
covariate(s) $X$, and a $k$-dimensional vector of parameters $\theta$ is minimized to obtain the 
parameter estimate $\hat{\theta}$ for the considered parametric model fitted in the current node of 
the tree. For this purpose many different M-type estimators can be applied including maximum likelihood 
or ordinary least squares. Therefore, a model-based discrepancy measure considered for the tree 
algorithm can be employed as objective function. 
To obtain additional information on the goodness of fit of the parametric model so~-~called scores are
calculated. Scores are the values of the derivative of the objective function with respect to the parameters 
with the estimated parameters and the observed values plugged in:
$$
S(y, x, \hat{\theta}) = \frac{\partial \psi(y, x, \hat{\theta})}{\partial \theta}
$$
which is a $n \times k$ - matrix for a $k$-dimensional parameter vector $\theta$ and
a data set of $n$ observations containing the response vector $y = (y_1,\ldots,y_n)^\top$
and regressor(s) $x = (x_1,\ldots,x_n)^\top$. 
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and parameter element $(y_i, x_i, \beta_l)$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$. 
After calculating these scores the aim is to find out whether there is any instability in the parameters
over each particular ordering induced by the possible partitioning variables. If a covariate has
no specific influence on the parameters the scores would always fluctuate randomly around their 
mean 0. Therefore, these deviations are investigated by measuring them using an empirical fluctuation 
process. Applying a scalar function which captures the fluctuation in this process yields the
test statistic of this so called generalized M-fluctuation test. To obtain the distribution of
the resulting test statistic asymptotic theory, in particular the central limit theorem is applied.
In this parametric unconditional framework p~-~values can be calculated and the covariate
with the smallest p~-~value is selected as split variable. Then an exhaustive search over all possible
partitions can be applied to find the best split point. Hence, the number of possible breakpoints is
not fixed unlike in the CTree algorithm which can make this operation computationally expensive.

\subsubsection{GUIDE}\label{sec:guide}
Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in the GUIDE algorithm.
From the model fitted to the data in the current node residuals are calculated. But before applying
a statistical test two additional steps are performed (to prepare(?) the data):
The residuals are binarized, i.e. depending on whether the sign of a residual is 
positive or not it is replaced by $1$ or $2$  % FIX ME: +1 and -1 in the implementation? 
and for each possible split variable the observed values are categorized into 4 groups 
separated by the corresponding 1st, 2nd and 3rd quartile.
Then, a $\chi^2$ test of independence is performed for the binarized residuals and
each categorized split variable.
After choosing the split variable showing the highest dependency by yielding the lowest p~-~value,
the split point minimizing the discrepancy measure is selected. The possible number of breakpoints
is determined by the categorization of the possible split variables such that the quartiles are
the only three possible split points.

\bigskip

\textit{Note:}
While GUIDE originally constructs large trees and applies post-pruning with the cross-validation 
method of CART, CTree and MOB employ pre-pruning by setting a significance level for the 
independence tests such that no split is performed if the lowest p~-~value is not smaller than 
the significance level.

\section{Unifying Framework}\label{sec:unify}
As all three algorithms follow the same basic structure their approaches can be presented together
in a unifying framework for a model based tree algorithm from which each specific algorithm can be 
obtained by switching on or off certain features. By applying such a general framework each of the 
characteristic features of the three algorithms and its impact can be investigated in detail. 
These switches are explained in this Section and their setting for the algorithms
CTree, MOB, and GUIDE are listed in Table~\ref{tab:switches}.

\subsubsection{Residuals vs full model scores}
Following the basic tree building strategy in Section~\ref{sec:urp}, a $n \times k$ - matrix is 
obtained based on the discrepancy measure. In the here discussed unifying framework this matrix 
is the score matrix $S$ as presented in Section~\ref{sec:mob} providing information on the 
goodness of fit for each observation and model parameter.
Since the first column of the score matrix is the vector of residuals the decision of using either 
residuals or full model scores for the test statistic is equivalent to the decision of using only 
the first column or the whole matrix $S$.

By including all available scores the goodness of fit of the model is investigated regarding each 
parameter rather than just regarding the mean/location parameter. In that way changes in parameters
other than the mean/location parameter should be detected more easily.

\textit{Example LM:} To exemplify this feature a linear model tree is considered as this allows 
for an easy and well structured description of the investigated subjects/issues. Therefore, from
here on this example will be used for illustrations.
(However, the conclusions drawn for this example also apply for a wide range of other model
types.)
The models fitted in the nodes of the considered linear tree model are of the form
$$
Y = \beta_0 + \beta_1 \cdot X + \epsilon
$$
for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
The coefficient vector $\boldsymbol{\beta} = (\beta_0, \beta_1)^\top$ is estimated
via minimizing the residual sum of squares (RSS) for the learning data set in 
the current node, for notational simplicity generally denoted by 
$\{(y_i, x_i)\}_{i=1\dots n}$. Hence, the discrepancy measure, i.e. the objective function is
$$
RSS(y_i, x_i, \boldsymbol{\beta}) = \sum_{i=1}^n (y_i - \beta_0 - \beta_1 \cdot x_i)^2
$$ 
such that the corresponding score matrix is defined by
$$
s(y_i, x_i, \hat{\boldsymbol{\beta}}) = \frac{\partial \textit{RSS}}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\boldsymbol{\beta}})
$$
which is in this case a $n \times 2$ - matrix as $\boldsymbol{\beta}$ is two dimensional. 
(In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and coefficient $(y_i,x_i,\beta_j)$ with $i \in \{1,\dots,n\}$
and $j \in \{1,2\}$.)
In this example considering only residuals corresponds to including the derivatives of $RSS$
only regarding the intercept parameter $\beta_0$, i.e. the first column of $S$.
On the contrary, using full model scores means including the whole $n \times 2$ matrix $S$, hence,
the derivatives of $RSS$ regarding the intercept $\beta_0$ (first column) and the slope parameter 
$\beta_1$ (second column).

\subsubsection{Binarization of scores}
Rather than applying the independence test directly on the selected (columns of) scores a
binarized version of them can be used as well. In particular, the score values are binarized
at 0 such that only their signs are considered.

\subsubsection{Categorization of covariates}
Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd and 
3rd quartile are used to bin each possible split variable into four categories.

%\subsubsection{Conditional vs unconditional}

\begin{table}[t!]
\centering
\caption[Table caption text]{CTree, MOB, and GUIDE with the corresponding setting of
the switches in the unifying framework and the approximation type of the underlying null distribution.}
\label{tab:switches}
\begin{tabular}{ l c c c c }
\hline 
& Scores            & Binarization  & Categorization  & Null distribution \\ 
\hline
CTree & full model scores & off           & off             & conditional\\
MOB   & full model scores & off           & off             & unconditional\\
GUIDE & residuals         & on            & on              & unconditional\\
\hline
\end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%All three algorithms apply statistical tests yielding p~-~values based on which
%the most appropriate split variable is chosen. One main difference lies in
%the way how scores are included. While CTree and MOB directly include the values
%of all available scores the GUIDE algorithm only uses residuals, in particular,
%the signs of residuals and binned versions of the covariates.
%The advantages of including all scores or only residuals as well as the impact
%of the binarization of scores at zero and the binning of covariates is
%investigated in detail in Section~\ref{sec:results}.
%Hence, the GUIDE algorithm differs mainly in the setting of the applied test
%in which the MOB and CTree algorithms are very similar. However, one essential difference
%between the testing strategies of MOB and CTree lies in the underlying approximation 
%of the null distribution where MOB employs an unconditional approach contrary
%to the conditional framework of CTree.

%Once the split variable is found they choose the split point minimizing the discrepancy measure, hence, 
%the RSS in this case.

%Once the split variable is chosen the search area for the optimal split point 
%is reduced to the range of this split variable. A possible but computationally 
%demanding approach to finding the best split point is an exhaustive search over 
%all possible split points in the split variable. However, other
%strategies have been developed as well, for example based on two-sample statistics such as 
%applied in the CTree algorithm.

\section{Simulation Setting and Evaluation}
\label{sec:simulation}

Two different scenarios are considered in this simulation study. First, the underlying
tree structure based on which the data is generated is a stump, i.e. a tree with only one 
split. By keeping the complexity of the model as low as possible the features/advantages and
disadvantages of the different testing strategies can be elaborated more precisely and 
the focus can be put on their power in terms of selecting the correct split variable.

In the second scenario the 'true' tree structure contains two splits in two different variables
yielding a tree with three terminal nodes. After having investigated the characteristics of 
the employed algorithms in detail for the stump scenario this additional evaluation aims at
showing that the same conclusions can be drawn for more complex structures.

In the following the data generating process is explained in detail for the first scenario.
The second, more complex scenario employs the same basic structure simply adding another split
as illustrated in Figure~\ref{fig:true_tree}.

\subsection{Data generating process}
Each generated data set consists of a response variable, one regressor, one split 
variable and nine noise variables. In particular, the variables listed in 
Table~\ref{tab:variables} together with the corresponding distributions are 
included in the data generating process.

\begin{table}
\begin{center}
\caption[Table caption text]{Variables included in the data generating process.}
\label{tab:variables}
\begin{tabular}{l l l}
Name                    & Notation                   & Specification \\
\hline
response                & $Y$                        & $\mathcal{N}(\mu(X, Z_1), \sigma^2)$ \\
mean of $Y$             & $\mu$                      & linear function of $X$ and $Z_1$ \\
variance of $Y$         & $\sigma^2$                 & 1\\
regressor               & $X$                        & $\mathcal{U}([-1,1])$ \\ %or $(-1)^{\mathcal{B}(0.5)}$ \\
true split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
coefficients            & $\beta_0$, $\beta_1$       & functions of $Z_1$\\
true split point        & $\xi$                      & $\in \{0, 0.2, 0.5, 0.8\}$\\
effect size             & $\delta$                   & $\in \{0, 0.25, 0.5, 0.75, 1\}$
\end{tabular}
\end{center}
\end{table}

The location parameter $\mu$ of the normally distributed response variable $Y$ depends
linearly on the regressor variable $X$ which is a uniformly distributed variable taking 
values between $-1$ and $1$.
% can either be a binary variable (taking either $0$ or $1$) or a uniformly distributed 
% variable (taking values between $-1$ and $1$).
Moreover, the intercept $\beta_0$ and the slope parameter $\beta_1$ of this linear equation
(can) depend on the true split variable $Z_1$ in the following way:
\begin{center}
$\mu(X, Z_1) = \beta_0(Z_1) + \beta_1(Z_1) \cdot X$.
\end{center}

In particular, three different cases are considered for the coefficients $\beta_0$ 
and $\beta_1$:
\begin{enumerate}
\item The intercept $\beta_0$ varies depending on $Z_1$ while $\beta_1$ is fixed (at $1$).
\item The slope coefficient $\beta_1$ varies depending on $Z_1$ while $\beta_0$ is fixed 
(at $0$).
\item Both coefficients $\beta_0$ and $\beta_1$ vary depending on $Z_1$.
\end{enumerate}

The varying coefficient(s) $\beta_i$, $i\in\{0,1\}$, are binary taking either the positive or
the negative value of the effectsize, i.e. $\delta$ or $-\delta$.
% can either be binary (taking either $\delta$ or $-\delta$), or a continuous function. 
% In the binary case the following definition is applied for the true split point $\xi$
\begin{align*}
\beta_{i}(Z_1) = \begin{cases}
-\delta \cdot (-1)^{i}  \quad \text{if } Z_1 < \xi \\
+\delta \cdot (-1)^{i}  \quad \text{if } Z_1 \geq \xi
\end{cases}
\end{align*}
%while in the continuous case a simple linear function is considered in the form of
%$$\beta_{i}(Z_1) = Z_1 \cdot \delta \cdot (-1)^{i}.$$
In that way, the type of variation is the same for $\beta_0$ and $\beta_1$, however, in 
opposite directions.
An illustration of the three different scenarios 
%for binary coefficients and a continuous regressor 
can be found in Figure~\ref{fig:dgp}.

%\newpage
\begin{figure}%[h!]
\setkeys{Gin}{width=1.0\linewidth}
<<dgp_bin_beta_cont_reg, fig=TRUE, echo=FALSE>>=
set.seed(7)

nobs <- 250
sigma <- 0.5
xi <- 0
delta <- 1
beta0 <- 0
beta1 <- 1

@@

par(mar = c(4,4,2,1))
plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("vary"~beta[0]),
     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
legend("top", c(expression(z[1]<0), expression(z[1]>=0)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
plot(x = d$z1, y = d$beta0, ylab = expression(beta[0]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1)) 
plot(x = d$z1, y = d$beta1, ylab = expression(beta[1]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1))

@@

par(mar = c(4,4,2,1))
plot(x = d$x[d$z1<xi], y = d$y[d$z1<xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("vary"~beta[1]),
     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
points(x = d$x[d$z1>=xi], y = d$y[d$z1>=xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
legend("top", c(expression(z[1]<0), expression(z[1]>=0)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
plot(x = d$z1, y = d$beta0, ylab = expression(beta[0]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1)) 
plot(x = d$z1, y = d$beta1, ylab = expression(beta[1]), xlab = expression(z[1]), pch = 21, bg = "black", cex = 0.7, ylim = c(-1,1))
